Simulated Observations of Young Protoplanetary Discs

### 2013-01-22 paper.tex
text changes

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionize this field, so it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ envelope with a structure similar to the pre-stellar core L1544. We focus on simple species and conclude that OCS is among the best tracers to measure the disc structure and kinematics. 
\end{abstract}

@@

OCS traces only the innermost region of the disc and can be used to examine the central regions. As OCS is not seen in outflows it can be used to view rotation of the central part of the disc without risk of contamination. As the excitation energy for the high J transitions of OCS is much higher than the other lines under consideration it is only excited in the hottest densest regions. As a result of this the spiral structure is clearer in OCS than in most transitions, with the low density regions between spiral arms showing no emission or absorption(see figure \ref{ocs_all}). The majority of the OCS is visible in absorption against the bright continuum emission from the disc midplane. The exception to this in the centre of the disc and is an effect of the inclination angle, whereby the line of sight passes through gas above the midplane and then through the hole in the centre of the disc.

@@

The abundance H$_2$CO molecule is fairly typical of the molecules simulated in Ilee (2011) in that it traces the spiral structure roughly the inner 2/3$^{rd}s$ of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation of detected out to larger radii than is possible with the OCS and is more visible between spiral arms but is otherwise simimlar in morphology.

@@

In order to make predictions on the observability of the synthetic brightness maps created by LIME, CASA (common astronomy software applications) was used to simulate observing these with the completed ALMA in configuration 26 of 28 for a total time of four hours, giving a beam size of between 0.02 and 0.03 arcseconds. In order to clean the images which featured mainly absorption the sky model used was the output of the lime simulations with the continuum subtracted and then the image inverted. The simulated images were cleaned down to a 3\sigma level with \sigma given by the online alma sensitivity calculator (ref).

@@

 \caption{C$^{17}$O 3-2 30 deg integrated intensity simulated for ALMA most extended configuration, contours starting at 3$\sigma$ and increasing in intervals of 3$\sigma$}
\end{figure*}

@@

 \caption{PVs with model rotation curve}
\end{figure*}

@@

 \caption{Spectra, H$_2$CO 20AU split, C$^{17}$O 28AU split, OCS 10AU split, HCO$^+$ 38AU split}
\end{figure*}

@@

In this paper we have performed radiative transfer simulations of a hybrid model comprising a 0.4$\, M_\odot$ self gravitating disc with radius 64$\,AU$ showing spiral density waves, surrounded by an envelope simulated as a collapsing 10$\,M_\odot$ BE-sphere. CASA simulations show that at a distance of 100$\,$pc both spatial resolution of the spirals in such a disc and extraction of kinematic and temperature from molecular lines are possible in ALMA band 7. Our simulations show that many  molecular species are predominantly seen in absorption towards the centre of  self-gravitating protoplanetary discs. The quiescent nature of the envelope around such discs only obscures them within $\pm\,0.5 km\,s^{-1}$.\newline
One assumption made in this model is that the gas and dust are in thermal equilibrium.If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption.\newline
%The mass of the disc used (0.4$\,\rm{M}_\odot$) is on the high end for early discs (citations for estimate of early class 1 disc masses).\newline
In the model spirals shock heat the mid-plane layer and the colder more diffuse gas absorbs the continuum from it, this results in molecular lines being seen in absorbtion whenever lines of sight pass through the cold diffuse gas at larger heights on to the hot dense midplane of the disc. Rotation curves can be gathered from these observations even with spiral structure\newline
Outflows could contaminate measurements of rotation curves meaning that speices  which commonly trace outflows, such as CO and HCO$^+$ are not going to be good tracers of disc rotation. However speices such as OCS and C$^{18}$O which are not commonly seen in outflows (e.g. van der Tak et al. 2003, Stanke et al. 2007, Yildiz et al 2012, Ren et al. 2011)can be used to trace disc rotation.

@@

\section[]{Other Inclinations}

>>>>

\begin{abstract}
The formation and earliest stages of protoplanetary discs are still lacking observational constraints. ALMA will soon revolutionise this field, so it is important to provide predictions and help the interpretation of future high sensitivity and high angular resolution observations. Here we present simulated ALMA observations of a relatively massive (0.39 M$_{\odot}$) self-gravitating disc embedded in a 10\,M$_{\odot}$ envelope with a structure similar to the pre-stellar core L1544. We focus on simple species and conclude that OCS is among the best tracers to measure the disc structure and kinematics. 
\end{abstract}

@@

OCS traces only the innermost region of the disc and can be used to examine the central regions. As OCS is not seen in outflows it can be used to view rotation of the central part of the disc without risk of contamination. As the excitation energy for the high J transitions of OCS is much higher than the other lines under consideration it is only excited in the hottest densest regions. As a result of this the spiral structure is clearer in OCS than in most transitions, with the low density regions between spiral arms showing no emission or absorption(see figure \ref{ocs_all}). The majority of the OCS is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this in the centre of the disc and is an effect of the inclination angle, whereby the line of sight passes through gas above the mid-plane and then through the hole in the centre of the disc.

@@

The abundance H$_2$CO molecule is fairly typical of the molecules simulated in Ilee (2011) in that it traces the spiral structure roughly the inner 2/3$^{rd}s$ of the disc. The spiral structure of the disc can be seen clearly in the integrated intensity map and its rotation of detected out to larger radii than is possible with the OCS and is more visible between spiral arms but is otherwise similar in morphology.

@@

In order to make predictions on the observability of the synthetic brightness maps created by LIME, CASA (common astronomy software applications) was used to simulate observing these with the completed ALMA in configuration 26 of 28 for a total time of six hours, giving a beam size of between 0.02 and 0.03 arcseconds. In order to clean the images which featured mainly absorption the sky model used was the output of the lime simulations with the continuum subtracted and then the image inverted. The simulated images were cleaned down to a 3\sigma level with \sigma given by the on-line ALMA sensitivity calculator (ref).

Sensitivities:
C17O J=$3\rightarrow2$ at 337.06GHz \sigma=0.0270$\,$mJy$\,$beam$^{-1}$
OCS J=$26\rightarrow25$ at 316.15GHz \sigma=0.0275$\,$mJy$\,$beam$^{-1}$
HCO$^+$ J=$3\rightarrow2$ at 267.56GHz \sigma=0.0176$\,$mJy$\,$beam$^{-1}$
H$_2$CO J=$4_{04}\rightarrow3_{03}$ at 290.62GHz \sigma=0.0228$\,$mJy$\,$beam$^{-1}$

@@

 \caption{Integrated intesity maps for the simulated observations of:{\bf Top Left:} H$_2$CO 4$_{04}$\rightarow3$_{03}$, {\bf Top Right:}C$^{17}$O 3\rightarrow2, {\bf Bottom Left:} OCS 26\rightarrow25 and {\bf Bottom Right:} HCO$^+$ 3\rightarrow2. contours start at 3$\sigma$ and increasing in intervals of 3$\sigma$ for each of the maps except the H$_2$CO which starts at 5\sigma and increases in intervals of 5\sigma}
\end{figure*}

The simulated integrated intesity maps demonstrate how different species allow observations to trace different regions of the disc, from OCS tracing the innermost regions to HCO$^+$ showing the outer regions which are less bright in the continuum.

@@

 \caption{Position velocity diagrams for each of the species simulated with CASA. The green curve on each diagram is the best fit curve to the azimuthally averaged rotation profile from figure 4.}
\end{figure*}

@@

 \caption{Spectra for each of the simulated species. In each diagram the three spectra are taken from positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=\pm N (plus blue, minus red) where N is different for each species. the values for N are: H$_2$CO 20AU, C$^{17}$O 28AU, OCS 10AU, HCO$^+$ 38AU. {\bf maybe should put stars or something on the mom0 maps to show these locations?}}
\end{figure*}

@@

In this paper we have performed radiative transfer simulations of a hybrid model comprising a 0.4$\, M_\odot$ self gravitating disc with radius 64$\,AU$ showing spiral density waves, surrounded by an envelope simulated as a collapsing 10$\,M_\odot$ BE-sphere. CASA simulations show that at a distance of 100$\,$pc both spatial resolution of the spirals in such a disc and extraction of kinematic and temperature from molecular lines are possible in ALMA band 7. Our simulations show that many  molecular species are predominantly seen in absorption towards the centre of  self-gravitating protoplanetary discs. The quiescent nature of the envelope around such discs only obscures them within $\pm\,0.5 km\,s^{-1}$.\newline
One assumption made in this model is that the gas and dust are in thermal equilibrium.If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption.\newline
%The mass of the disc used (0.4$\,\rm{M}_\odot$) is on the high end for early discs (citations for estimate of early class 1 disc masses).\newline
In the model spirals shock heat the mid-plane layer and the colder more diffuse gas absorbs the continuum from it, this results in molecular lines being seen in absorption whenever lines of sight pass through the cold diffuse gas at larger heights on to the hot dense mid-plane of the disc. Rotation curves can be gathered from these observations even with spiral structure\newline
Outflows could contaminate measurements of rotation curves meaning that species  which commonly trace outflows, such as CO and HCO$^+$ are not going to be good tracers of disc rotation. However species such as OCS and C$^{18}$O which are not commonly seen in outflows (e.g. van der Tak et al. 2003, Stanke et al. 2007, Yildiz et al 2012, Ren et al. 2011)can be used to trace disc rotation.

@@

\section{Other Inclinations} \label{sec:other_inc}
